---
title: "Week 3 NYPD Shootings Rmd"
author: "M. Aragon"
date: "2024-08-12"
output: html_document
---



```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)

```


# Week 3 NYPD Shooting Analysis

### Objective

While reviewing the data for this project, I'd like to answer the following questions:

  + What trend over time, if any, exists within the data? I'd like to understand when shootings generally occur;
  + What does the data tell me about each borough? Based on the shootings, where would be the safest place to visit based on the data?;
  + After answering the above, where would be the optimal place to visit and what time of year?


### How to retrieve the NYPD Shooting data

1. You'll want to add the tidyverse library to your session.  I've done this at the top of this rmd.

2. Obtain the correct URL for the file. Take a look at this and ensure it is what you need.  <https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD>

3. Call the read_csv import, passing it the URL from Step 2.  Place this into a variable.
  + I'm also creating a variable to pass to my visualizations to help with the presentation of each viz.
```{r}
shootings <- read_csv(file="https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")
breaksOverTheWholeRange = 12
```

### Review and transform the data

1. Review the data, outputting the results of the variable from the prior step.  In the data set, I can see:
  + What appears to be a unique value for each record in the data,
  + When the shooting took place,
  + Where the shooting took place (sometimes this data is missing),
  + What general values were captured regarding the shooting (perpetrator information as well as victim information), and,
  + Whether or not the shooting resulted in a murder.

```{r}
shootings
```
2. Tidy and Transform the data
  + Rename fields to make them easier to read/display
  + Modify the OccurrenceDate (from the prior step) to be a date field rather than a character field
  + Remove fields that we won't use 
  + Place null or NA values into the UNKNOWN perpetrator race category
  + Review the revised data set

```{r}
shootings <- shootings %>% 
rename(OccurrenceDate = OCCUR_DATE, OccurrenceTime = OCCUR_TIME, Borough = BORO, Location = LOCATION_DESC, Murder = STATISTICAL_MURDER_FLAG, PerpetratorRace = PERP_RACE, VictimRace = VIC_RACE, VictimSex = VIC_SEX) %>%
mutate(OccurrenceDate = mdy(OccurrenceDate)) %>%
mutate(PerpetratorRace = case_when(PerpetratorRace == "AMERICAN INDIAN/ALASKAN NATIVE" ~ "AMERICAN INDIAN/ALASKAN NATIVE",
                                  PerpetratorRace == "ASIAN / PACIFIC ISLANDER" ~ "ASIAN / PACIFIC ISLANDER",
                                  PerpetratorRace == "BLACK" ~ "BLACK",
                                  PerpetratorRace == "BLACK HISPANIC" ~ "BLACK HISPANIC",
                                  PerpetratorRace == "WHITE" ~ "WHITE",
                                  PerpetratorRace == "WHITE HISPANIC" ~ "WHITE HISPANIC",
                                  TRUE ~ "UNKNOWN")) %>%
select(-c(INCIDENT_KEY, LOC_OF_OCCUR_DESC, Lon_Lat, X_COORD_CD, Y_COORD_CD, JURISDICTION_CODE, LOC_CLASSFCTN_DESC, PERP_AGE_GROUP, Latitude, Longitude, Location, PERP_AGE_GROUP, PERP_SEX))

shootings
```
3. Make a new column of data to allow us to summarize Murders over time
```{r}
shootings$Murders <- with(shootings, ifelse(Murder == TRUE, 1, 0))
```
4. Do the same for shootings. Note: Everything in the file is a shooting, we just want a number to be able to summarize it.
```{r}
shootings$Shootings <- with(shootings, 1)
```
5. Add a column for the YearQuarter and Quarter and Year. Note: As I worked with the data, I tried many variables to see how the data would be best presented. So, yes, I am carrying some extra, unnecessary variables, which can be removed.
```{r}
shootings$YearQuarter <- with(shootings, quarter(OccurrenceDate, with_year = T))
shootings$Quarter <- with(shootings, quarter(OccurrenceDate, with_year = F))
shootings$Year <- with(shootings, year(OccurrenceDate))
shootings$YearMonth <-with(shootings, as.numeric(format(OccurrenceDate, '%Y%m')))
```
6. Assign color to the various perpetrator races.  Note: This is one variable that I could not get wired into the visualizations.
```{r}
shootings$PerpetratorRaceColor <- 
  with(shootings, case_when(
                        PerpetratorRace == "AMERICAN INDIAN/ALASKAN NATIVE" ~ "green",
                        PerpetratorRace == "ASIAN / PACIFIC ISLANDER" ~ "red",
                        PerpetratorRace == "BLACK" ~ "black",
                        PerpetratorRace == "BLACK HISPANIC" ~ "brown",
                        PerpetratorRace == "WHITE" ~ "white",
                        PerpetratorRace == "WHITE HISPANIC" ~ "tan",
                        PerpetratorRace == "UNKNOWN" ~ "lightgrey"))

```

7. Review the data
```{r}
shootings %>%
  select(OccurrenceDate, Murders, Shootings, YearQuarter, Quarter, Year, PerpetratorRace, PerpetratorRaceColor, YearMonth)
```
8. Summarize the data by quarter
```{r}
ShootingsOverTime <- shootings %>%
 group_by(YearQuarter) %>%
 summarize(Murders = sum(Murders), Shootings = sum(Shootings))
```

### Now, let's visualize the data and see what stands out

1. Display the data in a line chart. NOTE: To reduce the number of data points, I have summed up the data at the Year-Quarter level.
```{r fig.align='center', fig.width=12, fig.height=5}

ShootingsOverTime %>%
 filter(Shootings > 0) %>%
 ggplot(aes(x = YearQuarter, y = Shootings)) +
 geom_line(aes(color = "Shootings", alpha = 0.5)) +
 geom_point(aes(color = "Shootings")) + 
 geom_line(aes(y = Murders, color = "Murders", alpha = 0.5)) +
 geom_point(aes(y = Murders, color = "Murders")) +
 geom_smooth(method = "lm") +
 scale_color_manual(values = c("red", "blue")) +
 scale_x_continuous(n.breaks = breaksOverTheWholeRange) +
 labs(
     title = paste("Shootings and murders over time"), 
     x = "By Quarter & Year",
     y = "Number of shootings/murders"
 )
```

2. After reviewing the data, I noticed an interesting trend in shootings where Q2 and Q3 typically increase while Q4 and Q1 decrease. This raises the idea that people in the New York area must be outdoors more in the warmer months, leading to the higher number of shootings, versus the cooler/colder months of the year.  
  + After doing a Google search for the criteria "average monthly temperature in New York boroughs from 2006 to 2023", I came across this National Centers for Environmental Information [site](https://www.ncei.noaa.gov/access/monitoring/climate-at-a-glance/city/time-series). Within the site I was able to construct an average monthly temperature for New York's Central Park for the desired date range.
  + I did this by selecting the following paramters, reviewing the Plot feature to ensure it was what I wanted, then took a copy of the Download "CSV" button. Here are my parameters:
  + Parameter: Average Temperature
  + Time Scale: 1-Month
  + Month: All Months
  + Start Year: 2006
  + End Year: 2023
  + State: New York
  + City: New York (Central Park)
  
  + Next, I read the data in to a variable and reviewed the data. Note: This took a couple of tries as I had to figure out how many rows to skip to get to the values.
  + In addition, I renamed variables to make them easier to use and more meaningful.


```{r}
noaaAvgTemps <- read.csv(file="https://www.ncei.noaa.gov/access/monitoring/climate-at-a-glance/city/time-series/USW00094728/tavg/1/0/2006-2023.csv?base_prd=true&begbaseyear=2006&endbaseyear=2023", skip = 4)

noaaAvgTemps <- noaaAvgTemps %>%
  rename(YearMonth = Date, MonthlyAvgTemp = Value)

head(noaaAvgTemps, 10)
```

3. Bring the NOAA data in with the shooting data. Summarize the shootings by month, join the monthly average temperatures, then plot the data to see how shootings look compared to temperature.  NOTE: Because of the large number of months, I chose to narrow the date range to the last 8 years of data.

```{r fig.align='center', fig.width=12, fig.height=5}
# MonthlyShootingsWithTemp <- shootings
shootingsByMonth <- shootings %>%
  filter(Year >= 2016) %>%
  group_by(YearMonth, Year) %>%
  summarize(Shootings = sum(Shootings))

shootingsByMonth <- inner_join(shootingsByMonth, noaaAvgTemps, by = "YearMonth")
# shootingsByMonth
sortedTime <- sort(unique(shootingsByMonth$YearMonth), decreasing = FALSE)

# view(shootingsByMonth %>% filter(Year >= 2019))

for(yr in unique(shootingsByMonth$Year)){
  
  print(
  shootingsByMonth %>%
    filter(Year == yr) %>%
    ggplot(aes(factor(reorder(YearMonth, Shootings, sum), sortedTime), y = Shootings, alpha = 0.5)) +
    geom_bar(mapping = aes(x = YearMonth, y = Shootings), stat = "identity") +
    geom_line(mapping = aes(x = YearMonth, y = MonthlyAvgTemp)) +
    scale_x_continuous(
      breaks = scales::pretty_breaks(n = 12)
    ) +
    scale_y_continuous(
      expand = expansion(mult = c(0, .05)),
      sec.axis = sec_axis(
                    ~./1, 
                    name = "Avg Temperature", 
                    breaks = seq(
                        0, 110, by = 10))) +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(
      title = paste(yr, " shootings by all races"),
      x = "Month over month"
    )
  
  
  )
}

```

4. With the average monthly temperature overlayed on the monthly number of shootings, we can see that the summertime months are the peak time in the number of shootings over each of the years presented.

5. Let's see how the data looks for each Borough. 
  + Summarize the data by Borough
  + Then plot by Borough
  
```{r fig.align='center', fig.width=12, fig.height=5}
ShootingsOverTimeByBorough <- shootings %>%
  group_by(Borough, YearQuarter) %>%
  summarize(Murders = sum(Murders), Shootings = sum(Shootings))

for(boro in unique(ShootingsOverTimeByBorough$Borough)) {
  # print(boro)
  
  print(
    ShootingsOverTimeByBorough %>%
     filter(Shootings > 0 & Borough == boro) %>%
     ggplot(aes(x = YearQuarter, y = Shootings)) +
     geom_line(aes(color = "Shootings", alpha = 0.5)) +
     geom_point(aes(color = "Shootings")) + 
     geom_line(aes(y = Murders, color = "Murders", alpha = 0.5)) +
     geom_point(aes(y = Murders, color = "Murders")) +
     geom_smooth(method = "lm") +
     scale_color_manual(values = c("red", "blue")) + 
     scale_x_continuous(n.breaks = breaksOverTheWholeRange) +
     theme_minimal() +
     labs(
       title = paste(boro, " shootings and murders over time"), 
       x = "By Quarter & Year",
       y = "Number of shootings/murders")
    )
}

```  

6. Based on these charts, if I were to visit the New York area, I'd visit between the months of October and March, and travel to the safest location, which happens to be Staten Island (it has the fewest shootings and is trending downward across time).


### Next, let's visualize the data based on race over the different boroughs

1. Group up the data based on the Borough, Year, and Perpetrator Race. Then, plot the shootings over time.
```{r fig.align='center', fig.width=20, fig.height=10}
ShootingsOverTimeByBoroughandRace <- shootings %>%
  group_by(BoroughYear = paste(Borough, " ", Year), Borough, PerpetratorRace, PerpetratorRaceColor) %>%
  summarize(Murders = sum(Murders), Shootings = sum(Shootings))

TotalShootingOverTimeByBorough <- shootings %>%
  group_by(BoroughYear = paste(Borough, " ", Year)) %>%
  summarize(TotalShootings = sum(Shootings))

ShootingsOverTimeByBoroughandRace <- left_join(ShootingsOverTimeByBoroughandRace, TotalShootingOverTimeByBorough, by = "BoroughYear")

ShootingsOverTimeByBoroughandRace <- ShootingsOverTimeByBoroughandRace %>%
  mutate(PercentOfTotalShootings = (Shootings / TotalShootings)*100)


for(boro in unique(ShootingsOverTimeByBoroughandRace$Borough)){
  print(boro)
  
  boroData <- ShootingsOverTimeByBoroughandRace %>%
      filter(Shootings > 0 & Borough == boro)
  
  sortedTime <- sort(unique(boroData$BoroughYear), decreasing = FALSE)
  
  print(
        ggplot(boroData, aes(factor(reorder(BoroughYear, Shootings, sum), sortedTime), Shootings, fill = PerpetratorRace)) +
          geom_col() +
          geom_text(
            aes(label = after_stat(y), group = BoroughYear),
            stat = 'summary', fun = sum, vjust = -1
          ) +
          theme(axis.text.x = element_text(angle = 90)) +
          labs(
            title = paste(boro, " shootings by perpetrator race"),
            x = "Shootings over the years"
          ) + scale_fill_manual(values = c("#d6d2d2", "#f25757", "#faa381", "#fae588", "#b4ebca", "#83bcff", "#8a84e2"))
          
          
  )
   # scale_color_manual( values = c('red','blue','green'))
  
  # colors but mis-assigns ...
  # 
  
}
```

2. When reviewing the data based on the borough, over time, and broken out by the perpetrator race, it concerns me that there is such a large portion of shootings in every borough where the perpetrator information is either not collected/NA or UNKNOWN. Because of this, it makes it very difficult to accurately state which race may be the cause of the most shootings. However, if the UNKNOWN category is ignored the perpetrator race of BLACK appears to lead the number of shootings in every borough in every year of the data set.

3. Here's an alternate view of the data over time, excluding the UNKNOWN values
```{r fig.align='center', fig.width=20, fig.height=10}

shootings %>%
  group_by(Year, Borough, PerpetratorRace) %>%
  summarize(TotalShootings = sum(Shootings)) %>%
  
  ggplot(aes(Year, TotalShootings)) +
  geom_point(aes(colour = PerpetratorRace,
                 size = TotalShootings,
                 alpha = 0.5)) +
  facet_wrap(~Borough, nrow = 5) +
  theme_bw()
```

### Conclusions

While reviewing the NYPD Shootings data set, I attempted to plot the data as-is, with very little modification as well as personal bias. I did, however, group NA and (null) Perpetrator Races in with the UNKNOWN Perpetrator Race category given that, to me, UNKNOWN is UNKNOWN. In addition, I did a number of searches looking for why shootings may increase in New York and I kept coming back to the same thought - temperature/weather.  In doing so, I was able to find a government agency, National Centers for Environmental Information, which happened to store average temperatures for multiple New York areas over the span of our data set. By layering the temperature data in with the shooting data, it became very apparent that shootings in all of the boroughs of New York increase during the summertime months.

While I wanted to delve deeper into the perpetrator race, the UNKNOWN data makes this a very tricky topic. It is not fair to assume that all of the UNKNOWN perpetrators were black. Nor is it fair to assume they are White or Hispanic or any other race for that matter. So, the most I could do is simply report the facts: 

 * Shootings are related to weather in the New York area. Summertime appears to be the most active time for shootings.
 * Perpetrators with a race of Black appear to contribute to a large number (majority) of the shootings across the 5 boroughs but we cannot say definitively that they are THE race who cause the most shootings. In order to do that, we'd need to accurately determine the UNKNOWN perpetrator race.
 * Because of the trends in the data, I'll be sure to only visit the New York area between October and March and it may very well be (sadly) Staten Island.  I say "sadly" because visiting the Empire State building along with the skyscrapers of Manhattan has long been a dream of mine.
  
By sticking to just the facts, I believe I have kept my own personal bias out of my analysis as well as limited the bias of my source data, by only using the NYPD Shooting data set and data from a government agency.

